---
title: "The Wreck of the Titan Sentiment Analysis"
runningheader: "Sentiment Analysis" # only for pdf output
subtitle: "" # only for html output
author: "B.M Njuguna"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: biblatex
    latex_engine: xelatex
number_sections: no
toc: no
toc_depth: 4
nocite: '@'
bibliography: ref.bib
biblio-style: apa
link-citations: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(cache.extra = packageVersion('tufte'), comment = NA, warning = FALSE,
                      message = FALSE, cache = TRUE, echo = TRUE)
options(htmltools.dir.version = FALSE)

```

```{r, loading-packages,include=FALSE}
# Loading the packages
library(pacman)
p_load(readxl, dplyr, ModelMetrics, tm, gmodels, ggplot2, wordcloud, knitr)
```

\section{\textcolor{red}{\large\bfseries{Chapter One: Introduction}}}

\subsection{\textcolor{blue}{1.o Background Information}}

The **Wreck of the Titan** is a novel that was written by American
writer **Morgan Robertson** (@robertson1912wreck) . It was first
published in 1898, and then later republished in 1912. The fictional
story is about a ship called *Titan* which sinks in the North Atlantic
Ocean, after hitting a iceberg, resulting to loss of most of its
passengers and crew. This novel was written 14 yeas before the sinking
of the RMS Titanic ship in 1912. The similarities between the fictional
story and the actual event of the sinking of the ship in 1912 have led
to much speculations about whether Robertson's may have inspired or even
predicted the Titanic Disaster.

Robertson describes the ship as a large unsinkable passenger liner with
many of the features similar to the titanic (@donnelly2000titanic). The
ship in the fictional story travels at high speed in a sea filled with
icebergs which is similar to the Titanic ship. The book's description
about the sinking of the ship have similarities with the actual
description of the sinking of the ship. Also, in the book, there are
challenges such as lack of enough life boats and confusion among the
crew which are similar to the actual sinking of the ship. The books
prescient warning about the dangers of the overconfidence in the face of
natural disasters.

**Sentiment analysis** is a technique used to extract, identify and/or
identify subjective information from text or novel
(@mejova2009sentiment). It is a sub-field of Natural Language Processing
that is also used to identify whether the sentiment of a text, novel or
document is positive, negative or neutral. Sentiment analysis started in
the 1950s with simple use of lexical (lexicons)[^1] and heuristics
database to identify and classify words as positive and negative.
However, these early attempts were limited by their accuracy and
effectiveness.

[^1]: Lexicons are dictionaries or lists of words and their meanings,
    often used in natural language processing and computational
    linguistics. They can also include additional information about
    words such as part-of-speech, pronunciation, and syntactic behavior.
    Lexicons can be created manually or automatically generated from
    large text corpora.

The field of sentiment analysis advanced significantly at around 2000
with the invention of more sophisticated machine learning algorithms and
the availability of large amounts of labelled data for training and
testing (@mantyla2018evolution).

\subsection{\textcolor{blue}{Objectives of the Study}}

1.  The objective of this study was to examine the general sentiment of
    the novel *The wreck of the titanic*.

2.  Examine the change of emotions across the trajectory of the novel.

3.  Compare the results of different performances of the lexicons used.

\section{\textcolor{red}{Methodology}}

Basically, sentiment analysis method was used in this study. The study
was done using *R statistical package*. Specifically, lexicons in R were
used to obtain the sentiment of the novels.

First, The R package *gutenbergr* was used to download the novel in R.
Then the R package *tidytext* was used to clean and put the text in the
novel in the tidy format easier analysis. *tidytext* was also used to
download the lexicons for the sentiment analysis. The three lexicons
used for the analysis were as listed below;

-   **AFINN**

-   **Bing**

-   **NRC**

All three lexicons are based on unigrams, i.e., single words. These
lexicons contain many English words and the words are assigned scores
for positive/negative sentiment, and also possibly emotions like joy,
anger, sadness, and so forth. The **NRC** lexicon **categorizes words in
a binary fashion ("yes"/"no") into categories of positive**, **negative,
anger, anticipation, disgust, fear, joy, sadness, surprise, and trust**.
The **Bing** lexicon **categorizes words in a binary fashion into
positive and negative categories**. The **AFINN** lexicon **assigns
words with a score that runs between -5 and 5, with negative** **scores
indicating negative sentiment and positive scores indicating positive
sentiment**.

\section{\textcolor{red}{Data Analysis}}

The novel was downloaded in R and the glimpse of the first part of the
first page is shown below;

```{r}
# Checking which id is the novel
# Foirst we llok up for novels written by Robertson

library(gutenbergr)
robertson_books <- gutenberg_works(author == "Robertson, Morgan")

# Downloading the novel after identifyig the ID of the novel

titanic <- gutenberg_download(24880)
cat(head(titanic$text, c(100)), sep = "\n")
```
